---
title: "Extracting data from Facebook's Advertising Platform"
subtitle: "SOC6708 ADA"
author: "Monica Alexander"
format: 
  pdf:
    linkcolor: blue
number-sections: true
toc: true
execute: 
  warning: false
  message: false
---

# Overview

In this module we will be querying Facebook's Marketing API to extract estimates of the number of Facebook users that have certain characteristics, behaviors and interests. Unlike module 1, we will not be relying on an R package which has 'wrapper functions' around API calls, but rather constructing the queries to the API ourselves. The structure of this module is:

- Loading in the R packages we need
- Setting up Facebook API clients
- Extracting basic population estimates
- Extracting population estimates by age and sex
- Extracting more detailed demographic estimates 
- Extracting estimates by region

# Load in the R packages we need

The first step is to load in all packages we need. We are relying on `tidyverse` and `plyr` for data manipulation. The `httr` package allows us to query websites, and the `jsonlite` package helps to manipulate data from JSON files. 

```{r}
#| message: false
library(tidyverse)
library(plyr)
library(httr)
library(jsonlite)
library(here)
```

# Facebook set up

Just like in the first module, in order to make calls to the API, you will need to have a standard Facebook account. If you don't already, you can sign up here: https://www.facebook.com/.

## Creating an app through the Facebook developers website

We need to create an app. Go to https://developers.facebook.com/. You will need to create a Facebook developer's account. Then, click on My Apps. Then, click Create App, name it and choose Marketing API. You will eventually be led to the app dashboard. 

## Check version
Now that we've created an app, we need to get some info to input into our queries. To check the version, select Settings and then Advanced. 

Insert the version number into code below and run the code:
```{r}
#| eval: false
# insert your version number 
version <- "v25.0" 
```

## Obtain app token

We also need to get the app token, which is unique to your app. From the menu go to Tools then Graph API Explorer. Copy the User Access Token and paste it where the x's are below and run the code below. Note you will have to play around with permissions (it will need access to ads management and ads read).



```{r}
#| eval: false
token <- "EAAUsL2bZBL4ABQ6icUr6YoddA7Ijv7USvBgZAQCcn7EVsr9aBbrnZB7ycwgRaM6rQsA2qXpl5pQqBnbESh7sDEZBxQ8P0143EvyR4TuZCpfrZAb4Vjw58UXhuZCh8NMTj8QwkYcnRd769MxllPZBDvtZCmcglGHloXOibEDXGbs4ZALoPWtHe2CnpTwVzjRez5JBIL4oFTFRpkrwExd9LHnYLiDNNIIrp35YC8whxfoAZDZD"
```

## Obtain creation act

The final piece of information required is your unique creation act. The easiest way to get this info is to go to facebook.com/adsmanager. Then, the url will automatically update and the act number is in the URL. Copy this and paste it below where the x's are and run the code:


```{r}
#| eval: false
act <- "1466954328287520"
```

## Set credentials

We now want to put all our credentials (act number, version, and token) into the following string which will allow us to make calls to Facebook's Marketing API. 

```{r}
Credentials <- paste0('https://graph.facebook.com/',version,
                      '/act_',act,'/delivery_estimate?access_token=',token,
                      '&include_headers=false&method=get&
                      optimization_goal=REACH&pretty=0&suppress_http_code=1')
```


## Exercise

Change the parameters in the code in order to retrieve the number of women between 20 and 55 years old that live in Australia that are Facebook users.

```{r}
country <- '"AU"'
age_min <- 20
age_max <- 55
gender <- 2 

query <- paste0(Credentials,'&
                targeting_spec={
                "age_min":',age_min,',
                "age_max":',age_max,',
                "genders":[',gender,'],
                "geo_locations":{"countries":[',country,']},
                "facebook_positions":["feed","instant_article",
                "instream_video","marketplace"],
                "device_platforms":["mobile","desktop"],
                "publisher_platforms":["facebook","messenger"],
                "messenger_positions":["messenger_home"]}')


query <- str_replace_all(query, "[\n]", "")  |> 
          str_replace_all(" ", "")

query_val <- url(query) %>% fromJSON

query_val$data

```


# Extracting more detailed information based on different demographics, behaviors, and interests

In addition to basic demographic characteristics, you can also extract information based on other demographics, behaviors, or interests. The documentation [here](https://developers.facebook.com/docs/marketing-api/audiences/reference/advanced-targeting) provides further information.

## Looking up the available demographics

Broadly, there are three different categories: demographics, interests, and behaviors. The first step in extracting more detailed data is to find out all the possible variables we can choose from. To do this, we are going to query information about our version of the app, using the `GET` function from the `httr` package. 


```{r}
demographics_df <- GET(
  "https://graph.facebook.com/v25.0/search",
  query=list(
    type='adTargetingCategory',
    class='demographics',
    access_token=token,
    limit=2000
  ))  %>%
  content(as="text", encoding = "UTF-8") %>% 
  fromJSON %>%
  .[[1]]
```

If we take a look at the top of this data, we can see that there's all sorts of interesting information, based on birthdays, family or life events, etc. 

```{r}
head(demographics_df)
```

In order to include these variables in query, we need to pull out three pieces of information: the ID, type, and name. Let's get an estimate for people aged 30-40 in Canada who have recently moved. As before, we can set our age and location variables:

```{r}
age_min <- 30
age_max <- 40
country <- '"CA"'
```

We also need to pull out the information on recently moved:

```{r}
recently_moved_info <- demographics_df |> 
  filter(name == "Recently moved")
```

## Defining the `flexible_spec` object

Now the additional information is defined in a `flexible_spec` JSON object:

```{r}
rm_json <- paste0(',"flexible_spec":[{"',recently_moved_info$type,
                  '":[{"id":"',recently_moved_info$id,
                  '","name":"',recently_moved_info$name,
                  '"}]}]')
```

## Making the query

Now we want to add this into our query:

```{r}
query <- paste0(Credentials,'&
                targeting_spec={"age_min":',age_min,',
                "age_max":',age_max,
                rm_json,',
                "geo_locations":{"countries":[',country,']},
                "facebook_positions":["feed","instant_article",
                "instream_video","marketplace"],
                "device_platforms":["mobile","desktop"],
                "publisher_platforms":["facebook","messenger"],
                "messenger_positions":["messenger_home"]}')

# remove spaces
query <- str_replace_all(query, "[\n]", "")  |> 
          str_replace_all(" ", "")

# extract
query_val <- url(query) %>% fromJSON
query_val$data

```

## More than one group

Above we were just considering one age group. Let's extract information on the people who recently moved across multiple age groups and plot. Firstly, define the age groups:

```{r}
age_mins <- seq(15, 50, by = 5)
age_maxs <- seq(20, 55, by = 5)
```

Now we can repeat the code above, but loop through the age groups and store in a tibble:

```{r}
recently_moved <- tibble()

for(i in 1:length(age_mins)){
  query <- paste0(Credentials,'&
                targeting_spec={"age_min":',age_mins[i],',
                "age_max":',age_maxs[i],
                rm_json,',
                "geo_locations":{"countries":[',country,']},
                "facebook_positions":["feed","instant_article",
                "instream_video","marketplace"],
                "device_platforms":["mobile","desktop"],
                "publisher_platforms":["facebook","messenger"],
                "messenger_positions":["messenger_home"]}')

# remove spaces
query <- str_replace_all(query, "[\n]", "")  |> 
          str_replace_all(" ", "")

# extract
query_val <- url(query) %>% fromJSON
recently_moved <- bind_rows(recently_moved, query_val$data |> 
                              mutate(age_group = age_mins[i]))
}

query_val$data
```

Now we can plot the estimated mau_lower_bound users by age group:

```{r}
recently_moved |> 
  ggplot(aes(age_group, estimate_mau_lower_bound)) + 
  geom_bar(stat = "identity")+
  labs(title = "People in Canada on Facebook who have recently moved",
       y = "Estimated lower bound monthly users", x = "Age")
```

## Exercise

Find an estimate of the number of 30-50 year olds in Canada who work in Sales. 

```{r}
age_min <- 30
age_max <- 40
country <- '"CA"'

recently_moved_info <- demographics_df |> 
  filter(name == "Sales")

rm_json <- paste0(',"flexible_spec":[{"',recently_moved_info$type,
                  '":[{"id":"',recently_moved_info$id,
                  '","name":"',recently_moved_info$name,
                  '"}]}]') 


query <- paste0(Credentials,'&
                targeting_spec={"age_min":',age_min,',
                "age_max":',age_max,
                rm_json,',
                "geo_locations":{"countries":[',country,']},
                "facebook_positions":["feed","instant_article",
                "instream_video","marketplace"],
                "device_platforms":["mobile","desktop"],
                "publisher_platforms":["facebook","messenger"],
                "messenger_positions":["messenger_home"]}')

# remove spaces
query <- str_replace_all(query, "[\n]", "")  |> 
          str_replace_all(" ", "")

# extract
query_val <- url(query) %>% fromJSON
query_val$data
```


# Extracting population by region

As a final illustration, we are going to extract estimates for the number of Facebook users by Canadian province. Firstly, I'm going to list all the names of the provinces and territories:

```{r}
# Names of provinces and territories
prov_terr <- list(
  "Alberta", 
  "British Columbia",
  "Manitoba",
  "New Brunswick",
  "Newfoundland",
  "Northwest Territories", 
  "Nova Scotia", 
  "Nunavut",  
  "Ontario", 
  "Prince Edward Island",
  "Quebec",
  "Saskatchewan",
  "Yukon"
)
```


## Extract keys for regions

Before we query the Marketing API, we need to get the unique 'keys' for each province/territory. To do this, we are going to query the API meta data, in a similar way to what we did above. For example, to extract information about Alberta:


```{r}
  GET(
    "https://graph.facebook.com/v25.0/search",
    query = list(
      location_types = "region",
      type = "adgeolocation",
      q = prov_terr[1], 
      access_token = token, 
      limit = 3
    )
  ) %>% 
    content(as="text", encoding = "UTF-8") %>% 
    fromJSON %>% 
    .[[1]]
```
Let's make use of `dplyr`'s `ldply` function to get this information in a neat way for all provinces:


```{r}
id_df <- ldply(prov_terr, function(x) {
  GET(
    "https://graph.facebook.com/v25.0/search",
    query=list(
      location_types = "region",
      type = "adgeolocation",
      q = x, # use province name here
      access_token = token, 
      limit = 3
    )
  ) %>% 
    content(as="text", encoding = "UTF-8") %>% 
    fromJSON %>% 
    .[[1]]
})
```

Now the `id_df` data frame gives us the required information about the region keys for each province/territory.

```{r}
id_df
```

We can use this information to alter our query. As an example, let's extract information for all users in Alberta. Firstly define the key

```{r}
region_id <- id_df[1, "key"]
```

Now instead of having the country in the `geo_locations`, we have region:

```{r}
query <- paste0(Credentials,'&
                targeting_spec={
                "geo_locations":{"regions":[ {"key":"', region_id,' "}]},
                "facebook_positions":["feed","instant_article",
                "instream_video","marketplace"],
                "device_platforms":["mobile","desktop"],
                "publisher_platforms":["facebook","messenger"],
                "messenger_positions":["messenger_home"]}')

# remove spaces
query <- str_replace_all(query, "[\n]", "")  |> 
          str_replace_all(" ", "")

# extract
query_val <- url(query) %>% fromJSON
query_val$data
```
Let's loop over all regions, extract the data, add a column for region name, and save the data:

```{r}
#| eval: false

region_estimates <- tibble()
for(i in 1:nrow(id_df)){
  region_id <- id_df[i, "key"]
  query <- paste0(Credentials,'&
                targeting_spec={
                "geo_locations":{"regions":[ {"key":"', 
                region_id,' "}]},
                "facebook_positions":["feed","instant_article",
                "instream_video","marketplace"],
                "device_platforms":["mobile","desktop"],
                "publisher_platforms":["facebook","messenger"],
                "messenger_positions":["messenger_home"]}')
  
  # remove spaces
  query <- str_replace_all(query, "[\n]", "")  |> 
    str_replace_all(" ", "")
  
  # extract
  query_val <- url(query) %>% fromJSON
  this_region <- query_val$data
  this_region <- this_region |> 
    mutate(name = id_df[i, "name"])
  region_estimates <- bind_rows(region_estimates, this_region)
  Sys.sleep(3)
}

region_estimates <- region_estimates |> 
  select(name, estimate_dau:estimate_mau_upper_bound) |> 
  mutate(date_collected = Sys.Date())

write_rds(region_estimates, "/Users/liangqishen/Desktop/fb_ca_province.rds")
```

## Comparing Facebook populations to Canadian Census

It would be interesting to compare the Facebook population distribution to that from the Canadian census. First, let's load in population estimates from StatCan. These are for the last quarter in 2022. I downloaded the data from [here](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000901).

```{r}
can_pop <- read_csv("/Users/liangqishen/Desktop/canadian_population.csv")
```

Now we want to compute the proportion of the total Canadian population by province/territory:

```{r}
can_pop <- can_pop |> 
  mutate(prop = population/sum(population))
```


Let's load in the Facebook data we collected and similarly calculate the proportion of users by province:

```{r}
fb <- read_rds("/Users/liangqishen/Desktop/fb_ca_province.rds")
can_fb_province <- fb |> 
  select(name, estimate_dau) |> 
  mutate(prop = estimate_dau/sum(estimate_dau)) |> 
  dplyr::rename(province = name, n_fb= estimate_dau, prop_fb = prop)

```


Join the proportions together and have a look. 

```{r}
can_pop |> 
  left_join(can_fb_province) |> 
  select(-population, -n_fb) |> 
  mutate_at(.vars = vars(prop:prop_fb), funs(round(., 2))) |> 
  arrange(-prop)
```

We can also plot these proportions to better visualize differences in proportions. First, some tidyverse to get the data into a easy plotting format:

```{r}
props_long <- can_pop |> 
  left_join(can_fb_province) |> 
  select(-population, -n_fb) |> 
  mutate(province = fct_reorder(province, prop)) |> 
  pivot_longer(-province, names_to = "source", values_to = "prop") |> 
  mutate(source = ifelse(source == "prop", 
                         "StatCan", "Facebook"))
```

Now plot:


```{r}
ggplot(props_long, aes(province, prop, fill = source)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  coord_flip() + 
  labs(title = "Canadian population and Facebook users",
       x = NULL,
       y = "proportion of total") + 
  theme_bw() + 
  scale_fill_brewer(palette = "Set1")
```


## More granular geographies

We can go even more granular with the geographies, from cities, to neighbourhoods, to zip codes. Details on levels and definitions are given here: https://developers.facebook.com/docs/marketing-api/audiences/reference/targeting-search/.

For example, let's get the key and estimate of users in Toronto:

```{r}
  GET(
    "https://graph.facebook.com/v25.0/search",
    query = list(
      location_types = "city",
      type = "adgeolocation",
      q = "Toronto", 
      access_token = token, 
      limit = 3
    )
  ) %>% 
    content(as="text", encoding = "UTF-8") %>% 
    fromJSON %>% 
    .[[1]]
```

```{r}
query <- paste0(Credentials,'&
                targeting_spec={
                "geo_locations":{"cities":[ {"key":"', "296875",' "}]},
                "facebook_positions":["feed","instant_article",
                "instream_video","marketplace"],
                "device_platforms":["mobile","desktop"],
                "publisher_platforms":["facebook","messenger"],
                "messenger_positions":["messenger_home"]}')

# remove spaces
query <- str_replace_all(query, "[\n]", "")  |> 
          str_replace_all(" ", "")

# extract
query_val <- url(query) %>% fromJSON
query_val$data
```
# Exercises

1. Extract estimates for the number of men in Ontario. 


2. Repeat 1. but for women, and combine the results of both into the one data frame with a gender column.
3. Using the `GET` function, extract the types of behaviors that are available to extract. 
4. Based on the information from part 3., extract estimates of the number of small business owners in Canada. 
5. Extract information on the number of small business owners who have also recently moved. 




