---
title: "Week 2: Mortality I lab"
subtitle: "SOC6708 ADA"
author: "Liangqi (Cecilia) Shen"
format: pdf
execute: 
  warning: false
  message: false
---

# Read in data

```{r}
library(tidyverse)
library(here)
library(readxl)
library(janitor)
```

For this example we are using mortality in Canada by cause, 2012-2016. Available from [StatCan](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310039201). 

```{r}
d <- read_csv(here("/Users/liangqishen/Desktop/CAN_age_cod.csv"))
head(d)
```


# Create a life table

Let's create a life table using the mortality rates from all causes of death in 2016. Get the data we need:

```{r}
dl <- d %>% 
  filter(year==2016, cause=="Total, all causes of death  [A00-Y89]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(year, age, Mx)

head(dl)
```

We need to create columns:

- $n$
- $_na_x$
- $_nq_x$
- $_nd_x$
- $_nL_x$
- $T_x$
- $e_x$

```{r}
d_lt <- dl |> 
     mutate(n = case_when(
       age==0 ~ 1,
       age==1 ~ 4,
       TRUE ~ 5
     ),
       ax = case_when(
       age==0 ~ 0.07 + 1.7*Mx,
       age==1 ~ 1.5,
       age==90 ~ 1/Mx,
       TRUE ~ 2.5
     ),
      qx = n * Mx / (1 + (n - ax)* Mx), # probability of death
      px = 1 - qx, # probability of survive
      lx = lag(cumprod(px), default = 1), 
      dx = lx - lead(lx, default = 0), #ndx = lx-Lx+n
      Lx = n * lead(lx, default = 0) + (ax* dx),
      Tx = rev(cumsum(rev(Lx))),
      ex = Tx / lx
      ) 
```

What's the life expectancy at age 10 in 2016?
```{r}
dl_age10 <- d_lt %>% 
  filter(age==10) %>% 
  select(year, age, ex)

head(dl_age10)
```


## Exercise

Calculate the lifespan disparity for Canada in 2016.
```{r}
d_lt |>
  mutate (prob = dx*ex) |>
  summarise(lifespan_disparity=sum(prob))

```
```{r}
d_lt |>
  group_by(year) |>
  mutate (prob = dx*ex) |>
  summarise(lifespan_disparity=sum(prob))

```
# Calculate cause-deleted life expectancy

Now calculate life expectancy if all intentional injuries were deleted. Get the data we need:

```{r}
dls <- d %>% 
  filter(year==2016, cause=="Total, all causes of death  [A00-Y89]"|cause=="Intentional self-harm (suicide)  [X60-X84, Y87.0]") %>% 
  mutate(age = start_age, Mx = rate/100000) %>% 
  select(age, cause, Mx) %>% 
  mutate(cause = ifelse(cause=="Intentional self-harm (suicide)  [X60-X84, Y87.0]", "suicide", "total")) %>% 
  spread(cause, Mx) %>% 
  rename(Mx_i = suicide,
         Mx = total)

head(dls)
```

## Exercise: Cause-deleted life expectancy

You need to create the same columns as above, but with the cause-deleted versions ($-i$). Do this by first creating the ratio $R_x^{-i} = \frac{M_x^{-i}}{M_x}$, use this to get $_nq_x^{-i}$, and the rest is the same. 

What's the cause-deleted life expectancy at age 10? What's the implied life lost due to suicide?

```{r}
dls <- dls |>
  mutate (Rx_minusi = (Mx-Mx_i)/Mx)
```

```{r}
lt_cd <-dls |> 
     mutate(n = case_when(
       age==0 ~ 1,
       age==1 ~ 4,
       TRUE ~ 5
     ),
       ax = case_when(
       age==0 ~ 0.07 + 1.7*Mx,
       age==1 ~ 1.5,
       age==90 ~ 1/Mx,
       TRUE ~ 2.5
     ),
      qx = n * Mx / (1 + (n - ax)* Mx), # probability of death
     qx_minusi = qx*Rx_minusi,
      px = 1 - qx, # probability of survive
      lx = lag(cumprod(px), default = 1), 
      dx = lx - lead(lx, default = 0), #ndx = lx-Lx+n
      Lx = n * lead(lx, default = 0) + (ax* dx),
      Tx = rev(cumsum(rev(Lx))),
      ex = Tx / lx,
     #Cause deleted COLUMUNS
     px_minusi = 1 - qx_minusi, # probability of survive
      lx_minusi = lag(cumprod(px_minusi), default = 1), 
      dx_minusi = lx_minusi - lead(lx_minusi, default = 0), #ndx = lx-Lx+n
      Lx_minusi = n * lead(lx_minusi, default = 0) + (ax* dx_minusi),
      Tx_minusi = rev(cumsum(rev(Lx_minusi))),
      ex_minusi = Tx_minusi / lx_minusi
      ) 

print(lt_cd) 
```
```{r}
lt_cd_age10 <- lt_cd %>% 
  filter(age==10) %>% 
  select(age, ex_minusi)

head(lt_cd_age10)
```


#DO NOT DO!!!!

# Decomposition next week

